Optimal Transport Problems

### 2018-03-27 reports/progress-report/progress.tex
more theoretical results

\documentclass[11pt]{amsart}
\usepackage{../am221}
\linespread{1.2}
\title{Optimal Transport Problems}
\newcommand{\R}{\mathbb{R}}
\usepackage{bbm}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\E}{\mathbb{E}}
\usepackage{cleveref}
\author{Jiafeng Chen\and Francisco Rivera}
\date{\today}
\usepackage{hyperref}
\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=blue}
\renewcommand{\b}{\mathbf}

\theoremstyle{remark}
\newtheorem{rmk}{Remark}
\begin{document}
    \maketitle
    \section{Theoretical foundations}
    The following is a concise review of Chapters 2 and 3 in 
    \cite{peyre2017computational}.

@@

    Given optimal solutions to the primal and dual, $\b P$, $(\b f, \b g)$,
    respectively, \emph{complementary slackness} implies that \[
    P_{ij} (C_{ij} - f_i - g_j) = 0
    \]
    holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
    vice versa.

>>>>

\documentclass[11pt,reqno]{amsart}
\usepackage{../am221}
\linespread{1.2}
\usepackage[margin=1.5in]{geometry}
\title{Optimal Transport Problems}
\newcommand{\R}{\mathbb{R}}
\usepackage{bbm}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\E}{\mathbb{E}}
%\usepackage[margin=1in]{geometry}
\author{Jiafeng Chen\and Francisco Rivera}
\date{\today}
\usepackage{hyperref}
\usepackage[capitalise,nameinlink]{cleveref}

\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}
\renewcommand{\b}{\mathbf}

\theoremstyle{definition}
\newtheorem{prop}{Proposition}
\theoremstyle{remark}
\newtheorem{rmk}{Remark}
\begin{document}
    \maketitle
    \section{Theoretical foundations}
    The following is a concise review of selected topics in Chapters 2 and 3 in 
    \cite{peyre2017computational}.

@@

    Given optimal solutions to the primal and dual, $\b P$, $(\b f, \b g)$,
    respectively, \emph{complementary slackness} implies that \[
    P_{ij} (C_{ij} - f_i - g_j) = 0 \tag{Complementary slackness}
    \]
    holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
    vice versa.

    \subsection{Dual-ascent algorithms and Hungarian algorithm}

    Let $S \subset \{1,\ldots,m\}$ and $S' \subset \{1,\ldots,n\}.$ Let $\one_S$
    be a vector of zeros with ones at location prescribed by $S$. Given $\b f,
    \b g$ a feasible pair for the dual, call $(i,j)$ a balanced pair if the
    condition $f_i + g_j = C_{ij}$
    binds, and inactive otherwise. Then 
    \begin{prop}

        Let $(\tilde {\b f}, \tilde {\b g}) = (\b f, \b g) + \epsilon (\one_S,
        -\one_{S'})$. Then $(\tilde {\b f}, \tilde {\b g})$ is dual feasible for
        a sufficiently small $\epsilon$ if for all $i$, \[
        (i,j) \text{ is balanced } \implies \text{$j \in S'$}.
        \]
    \label{prop:feasible}
    \end{prop}
    \begin{proof}
        For all $i\in S$, consider \[
        \epsilon_i = \min_{j: (i,j) \text{ inactive}} C_{ij} - f_i - g_j
        \]
        Let $\epsilon = \min_i \epsilon_i$. If $(i,j)$ is inactive, then \[
        \tilde f_i + \tilde g_j \le f_i + \epsilon + g_j\le C_{ij},
        \]
        which continues to be inactive. If $(i,j)$ is balanced, then $\tilde
        f_i + \tilde g_j = f_i + g_j = C_{ij}$. In either case, $\tilde {\b f},
        \tilde {\b g}$ continues to be feasible.
    \end{proof}

    \begin{prop}
    \label{prop:dualascent}
        If a dual feasible solution $(\b f, \b g)$ is not optimal, then there
        exists $S,S'$ such that $(\tilde {\b f}, \tilde {\b g}) = (\b f, \b g) + \epsilon (\one_S,
        -\one_{S'})$ is feasible for small enough $\epsilon$ with a strictly
        better objective.
    \end{prop}

    \begin{proof}
        Let $\mathcal B$ be the set of balanced edges $(i,j)$, and consider the
        bipartite graph on $\{1,\ldots,m\} \sqcup \{1,\ldots,n\}$ with edges in
        $\mathcal B$. Let there be a source node $s$ with capacitated edges to
        $\{1,\ldots,m\}$ with capacity $a_i$, and a sink node $t$ analogously
        with capacities $b_j$. Consider the maximum flow $\b F$ on the network
        via Ford-Fulkerson. If the throughput of the flow is one, then let $P_
        {ij} = F_{ij}$, which generates a primal feasible solution $\b P$.
        Strong duality would imply that $\b P, (\b f, \b g)$ are both optimal.

        Thus we may assume that the throughput is less than one. We label the
        nodes reached from $s$ where $\b F$ does not saturate capacity, as well
        as the nodes that contribute flow to any labeled nodes. Store the
        labeled nodes in $Q$. Since the graph is bipartite, $Q$ can be split
        into $S,S'$. Note that if $(i,j)$ is balanced and $i\in S$, then $j\in
        S'$. By \cref{prop:feasible}, $(\tilde {\b f}, \tilde {\b g})$ so
        defined is feasible. We wish to show that \[
        \one_S^T \b a - \one_{S'}^T \b b > 0.
        \]
        We readily conclude that this is indeed the case by accounting for the
        flow on the network.
    \end{proof}

    \cref{prop:dualascent} provides a template for the celebrated Hungarian
    algorithm \cite{kuhn2010hungarian}. The Hungarian algorithm solves the
    Kantorovich problem in the case when both $\b a, \b b \propto \one$. In a
    probability context for the Wasserstein distance between empirical measures,
    if the measures $\mu, \nu$ are both $d$-dimensional with $n$ observations,
    the Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$ for\footnote{In the $d=1$ case, the optimal transport problem reduces to
    sorting.} $d =
    1$
    \cite{bernton2017inference}.

    \section{Application value}

    \cite{bernton2017inference} is an example of recent work applying the
    Wasserstein distance to statistics and machine learning. 
    % Restricting to a subclass of problems called \emph{matching problems}, where
    % $\b a, \b b \propto \one$.
